---
title: "A four-step study-aware transferability workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A four-step study-aware transferability workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(microTransfer)
set.seed(20260722)
```

## Why the property must be declared first

A feature can transfer in at least three different senses:

- **membership**: it remains prevalent in an unseen cohort;
- **response**: its association with an external variable replicates in an
  independent cohort;
- **prediction**: a model trained without a cohort retains performance in that
  cohort.

These are not interchangeable claims. `microTransfer` therefore starts with an
explicit declaration and matches each property to a different validation
design.

## Simulated multi-cohort data

The example contains four cohorts, two classes, a continuous environmental
variable, five truly stable members, several study-specific pooled candidates,
and a small response signal.

```{r simulate}
studies <- paste0("S", 1:4)
n_per_study <- 30
metadata <- data.frame(
  sample_id = paste0("sample_", seq_len(4 * n_per_study)),
  study = rep(studies, each = n_per_study),
  environment = rep(seq(-1.5, 1.5, length.out = n_per_study), 4),
  state = rep(rep(c("A", "B"), each = n_per_study / 2), 4),
  stringsAsFactors = FALSE
)

features <- paste0("taxon_", seq_len(60))
counts <- matrix(
  rpois(nrow(metadata) * length(features), lambda = 0.4),
  nrow = nrow(metadata),
  dimnames = list(metadata$sample_id, features)
)

# Stable membership in all cohorts.
counts[, 1:5] <- counts[, 1:5] + matrix(
  rpois(nrow(metadata) * 5, 8), nrow(metadata)
)

# Pooled-only candidates abundant in one cohort.
counts[metadata$study == "S1", 6:15] <-
  counts[metadata$study == "S1", 6:15] + 12

# Response candidates; only the first two remain concordant in S4.
for (j in 16:20) {
  active <- metadata$study != "S4" | j <= 17
  counts[active, j] <- counts[active, j] +
    pmax(0, round(5 + 3 * metadata$environment[active]))
}

# Features carrying a class signal.
counts[metadata$state == "B", 21:25] <-
  counts[metadata$state == "B", 21:25] + 5
```

## Step 1: declare the property

```{r declare}
membership_claim <- declare_transferability(
  property = "membership",
  study_col = "study",
  feature_level = "taxon"
)
membership_claim
```

## Step 2: discover without the held-out cohort

Candidate discovery is repeated inside every outer fold. This prevents the
held-out cohort from influencing feature selection.

```{r discover}
discover_membership(
  counts,
  metadata,
  sample_col = "sample_id",
  study_col = "study",
  heldout = "S4",
  prevalence = 0.50
)$candidates
```

## Step 3: match validation to the property

### Membership

```{r membership}
membership_fit <- run_transferability_workflow(
  membership_claim,
  counts,
  metadata,
  sample_col = "sample_id",
  prevalence = 0.50,
  n_null = 99,
  matched = TRUE,
  seed = 20260722
)
membership_fit$summary
membership_fit$stable_features

pooled_membership_contrast(
  counts,
  metadata,
  sample_col = "sample_id",
  study_col = "study",
  thresholds = c(0.10, 0.30, 0.50)
)
```

### Environmental response

The first three cohorts define candidates. S4 is untouched until independent
projection.

```{r response}
response_claim <- declare_transferability(
  property = "response",
  study_col = "study",
  outcome_col = "environment"
)

response_fit <- run_transferability_workflow(
  response_claim,
  counts,
  metadata,
  sample_col = "sample_id",
  discovery_studies = c("S1", "S2", "S3"),
  validation_studies = "S4",
  min_prevalence = 0.05,
  min_total = 5,
  alpha = 0.20
)
response_fit$summary
```

### Prediction

The example uses a nearest-centroid classifier so the vignette has no modeling
dependency. A publication analysis can provide `glmnet`, random-forest, or
other `fit_fun` and `predict_fun` functions with the same interfaces.

```{r prediction}
prediction_claim <- declare_transferability(
  property = "prediction",
  study_col = "study",
  outcome_col = "state"
)

prediction_fit <- run_transferability_workflow(
  prediction_claim,
  counts,
  metadata,
  sample_col = "sample_id",
  features = features[21:25]
)
prediction_fit$summary

prediction_null <- prediction_permutation_test(
  counts,
  metadata,
  sample_col = "sample_id",
  study_col = "study",
  outcome_col = "state",
  features = features[21:25],
  permutations = 19,
  seed = 20260722
)
prediction_null$balanced_accuracy_p
```

## Step 4: audit sensitivity and classify conservatively

```{r sensitivity}
classify_transferability(membership_fit)
classify_transferability(response_fit)
classify_transferability(prediction_fit)

noise_fit <- label_noise_sensitivity(
  counts,
  metadata,
  sample_col = "sample_id",
  study_col = "study",
  outcome_col = "state",
  features = features[21:25],
  noise = c(0, 0.10, 0.20),
  replicates = 20,
  seed = 20260722
)
noise_fit$summary

membership_grid <- membership_sensitivity(
  counts,
  metadata,
  sample_col = "sample_id",
  study_col = "study",
  prevalence = c(0.30, 0.50, 0.70),
  min_depth = c(0, 50),
  min_total = c(0, 10),
  n_null = 0
)
head(membership_grid)
```

The evidence label is deliberately conservative and should accompany, not
replace, the fold-level effect sizes, uncertainty intervals, empirical nulls,
and any design limitations.
